---
layout: post
title: My First Blog Post
author: Nana LaRue
---

This is simply a template for the NRS529 Blog Posts.  

When you start you post all you need to do is:

1. Open the template file
2. Save this with the following naming convention: `YEAR-MO-DY-LastName.Rmd`
3. Edit the `YAML`.
4. Write your post
5. Save the `.Rmd` in postsRMarkdown
6. Commit changes (you can use the `Git` tab in RStudio).
7. Push those changes back up to your repository on GitHub.
8. Submit a pull request.

To make it a bit easier, here are some of the things mentioned on the first day of class.

This example with echo and evaluate.

```{r evalAndEcho}
x<-rnorm(100,10,5)
y<-jitter(x,1000)
z<-vector("character",100)
z[x>10]<-"High"
z[x<10]<-"Low"
df<-data.frame(x,y,z)
df
```

And an example that just echo's the code without evaluation:

```{r noEval, eval=FALSE}
# This is what a comment looks like
print("Some stuff to print (but not really, becuase this isn't evaluated!")
```

Here's a plot, without echoing :

```{r myPlot, echo=FALSE, fig.width=7, fig.height=6}
with(df,plot(x,y,col=factor(z)))
```

And to get more on this, check out [the first class blog post](http://scicomp2014.edc.uri.edu/posts/2014-01-27-Hollister.html)

